Sampling Bayesian probabilities given only sampled priors

### 2025-02-13 main.tex
some proofing

Consider an inference in which we have a vector of observable summary
statistics \vecc\ that we are using to constrain a set of parameters of
interest \vecq.  There is a model $\hatc(\vecq,\vecn)$ for the
observables which involves the parameters of interest, but also a
vector \vecn\ of nuisance parameters.  
We wish to characterized the Bayesian posterior probability
\begin{equation}
  p(\vecq | \vecc) \propto \int dn\, \likeli(\vecc | \vecq, \vecn) p(\vecq) p(\vecn),
\label{eq:posterior}
\end{equation}
where $\likeli(\vecc | \vecq, \vecn)$ is a known likelihood function
of the data, and $p(\vecq)$ and $p(\vecn)$ are priors on the
parameters.  This posterior is complex enough that it requires
approximation by the output of a Markov chain (MC) wandering across the space $(\vecq,\vecn).$

@@

A concrete example, which motivated this development,
is when the observable data \vecc\ are the binned 2-point correlation functions of
cosmic fields derived from a catalog of galaxies; the parameters of
interest are cosmological quantities such as $\Omega_m, \sigma_8,$
etc.; and the nuisance parameters $\vecn$
include the coefficients of some linear expansion of the redshift distributions $n(z)$ of the galaxies being observed:
\begin{equation}
  n(z) = \sum_{k=1}^{M} n_k b_k(z).
  \label{eq:nzbasis}
\end{equation}
The $b_k$ are a set of predetermined basis functions for the redshift
distribution.  In our case there are 10 distinct galaxy populations to be characterized, leading to $O(1000)$ parameters $n_k$ to be considered.

@@

Facing this problem for the cosmological analyses of the 3-year data (Y3)
of the \textit{Dark Energy Survey} (DES),
\citet{hyperrank} devised a scheme whereby
the samples of \vecn\ are placed in a grid within some $M$-dimensional
unit hypercube $\mathcal{H}$.  The coordinates \vecu\ within the hypercube
are considered the compressed parameters of $n(z),$ and the
decompression function
$\hat{\vecn}(\vecu)$ outputs the $\vecn_\alpha$ sample at the nearest grid point to
  \vecu.  This solves the problem of creating a continuous \vecu\
  domain, and maintains the equal prior probability of each \vecn\
  sample, but the function \emph{output,} and the resultant likelihood
  function of \vecu, are discontinuous.
Various strategies are proposed to assign the $\vecn_\alpha$ to the grid points in
$\mathcal{H}$ based on summary statistics, to reduce the
discontinuities.  But the function is never smooth.
As a consequence, many MC samplers become quite inefficient in
sampling of the cosmological posterior.  In particular, samplers such as \textsc{MultiNest} that assume continuity are rendered nearly non-functional.  As a result, the Y3 cosmological priors could not be evaluated with this method.  Instead, the \vecn\ samples were not used, and an \textit{ad hoc} $p(\vecn)$ was adopted which allowed only shifts and dilations of the mean $n(z)$ of the \vecn\ samples.

@@

We describe here an approach that is algebraically similar to that of \citet{hans}, but does not require 2nd condition of Gaussianity for the nuisance prior.  Our approach is to seek a linear compression of \vecn\ into a lower-dimensional set of parameters \vecu\ that projects away variations in \vecn\ that do not influence the likelihood $\likeli.$  Standard density estimators can then be applied to the \vecu\ values implied by the known \vecn\ samples to yield a prior $p(\vecu)$ that can be used for the MC chain of the cosmological posterior.  The model \hatc, and hence $\likeli,$ will be continuous over this low-dimensional \vecu\ space, and marginalization over \vecu\ will yield posterior probabilities very close to marginalization over the original \vecn.

@@

Since $\matX$ is idempotent, we can write
\begin{align}
  \matX & = V_X \proj_M V_X^T, \\
  \matY \equiv \matI-\matX & = V_X \proj_{-M} V_X^T,
\end{align}
where $V_X$ is unitary and the projection matrix $\proj_M$ is defined as
\begin{equation}
  \left(\proj_M\right)_{ij} \equiv
\begin{cases}
                                            1,  &  i=j\le M \\
                                            0,  & \text{otherwise}
\end{cases}
\end{equation}
and $\proj_{-M}=\matI-\proj_M.$  For a chosen rank $M$ of the
transformation matrix $X$, our task becomes to identify the
eigenvectors $V_X$ that minimize
\begin{align}
  \left\langle \chi^2\right\rangle & = \frac{1}{N_\alpha} \sum_\alpha
  \left[ \matY \vecn_\alpha\right]^T \matF  \left[\matY
                                     \vecn_\alpha\right] \\
       & = \trace \left[ \covm_n V_X \proj_{-M} V_X^T \matF V_X
         \proj_{-M} V_X^T \right], \\
  C_n & \equiv \left\langle \vecn \vecn^T \right\rangle.
\end{align}
This optimization is easier if we transform first the systematic variables to
$\vecn^\prime =\matT\vecn$ such that $\covm_{n^\prime}=\matI$, \ie\
make the elements of $\vecn$ uncorrelated and unit-variance.  This
is accomplished by finding the eigensystem $\covm_n=\matV_n \matLam_n
\matV_n^T$ and setting $\matT = \matLam_n^{-1/2} \matV_n^T$.  With
this transformation, we are now seeking a different unitary matrix
$\matV_{X^\prime}$ that minimizes
\begin{align}
  \left\langle \chi^2\right\rangle & = \trace \left[ \matI
    V_{X^\prime} \proj_{-M} V_{X^\prime}^T \left[ (T^{-1})^T \matF
      T^{-1} \right] V_{X^\prime} 
    \proj_{-M} V_{X^\prime}^T \right] \\
  & = \trace \left[ \proj_{-M} V_{X^\prime}^T \matV_G \matLam_G \matV_G^T
    V_{X^\prime} \proj_{-M} \right],
  \label{eq:mintr}\\
  \matG \equiv \left(T^{-1}\right)^T \matF   T^{-1} & = \matLam_n^{1/2} \matV_n^T
  \matF \matV_n \matLam_n^{1/2} = \matV_G \matLam_G \matV_G^T.
  \label{eq:defG}
\end{align}
The last line defines a transformed Fisher matrix $\matG$ and its
(non-negative) eigensystem.  Examination of \eqq{eq:mintr} reveals that this quantity
is minimized if the elements of the unitary matrices satisfy
$\matV_{X^\prime}^T \matV_G = \matI \; \Rightarrow \; \matV_{X^\prime} = \matV_G,$ and the eigensystem of $G$ is placed
in order of decreasing eigenvalue $\lambda^G_i.$ The
elements surviving the project functions yield
\begin{equation}
  \left\langle \chi^2\right\rangle = \sum_{i>M} \lambda_i^G.
  \label{eq:chiresid}
\end{equation}
In other words each eigenvalue of the matrix $\matG$ in \eqq{eq:defG}
gives the contribution to $\left\langle\chi^2\right\rangle$ of one
projection (mode) of $\vecn.$

@@

The orange violins in Figure~\ref{fig:violins} show the distributions
of the $n(z)$ values that are generated by drawing $\widetilde{\vecu}$
values from a unit Gaussian.  It is clear now that projecting away the
unobservable fluctuations in $n(z)$ has substantially reduced the
variance of the function.  To check whether we have achieved our goal
of leaving the \emph{observable} consequences of the $n(z)$ variation
unchanged, we calculate the distribution of
\begin{equation}
  \chi^2 =  \left[\hatc(\vecq,\vecn) - \hatc(\vecq,\vecn_0) \right]^T
                                            \left[\covm_c^{-1} \hatc(\vecq,\vecn) - \hatc(\vecq,\vecn_0)\right]
\label{eq:chihist}
\end{equation}
for the cases (1) the $\vecn$ are drawn from the input samples, and
(2) are generated using the procedure defined above.  This $\chi^2$
measures the deviation of the model from that implied by the mean
vector $\vecn_0$ at some chosen nominal value of cosmological
parameters $\vecq$ (we also fix the other nuisance parameters of the
DES model for this test).

Figure~\ref{fig:chihist} shows the results: the $\chi^2$ distribution
for the input samples is indistinguishable from that of the samples
generated by our compressed, normalized representation.
The Figure also shows the $\chi^2$ distribution resulting from samples
generated by the method used in the DES Y3 analysis \citep{y3pz}.  In
that case, the $n(z)$ for each galaxy sample was given an \textit{ad
  hoc} variation of the form
\begin{equation}
  n(z) = n_0\left(\frac{z-\Delta\,z}{s}\right),
\end{equation}
with $\Delta z$ and $s$ being ``shift'' and ``stretch'' parameters.
Separable Gaussian priors were assigned to $\Delta z$ and $s$ with
means of 0 and 1, respectively, and standard deviations that equaled
the RMS variation in the mean and width of the input $n(z)$ samples.
In the Figure we can see that the shift-stretch model does not in fact
reproduce the size of the deviations in the $\vecc$ observables that
is implied by the original samples.

>>>>

Consider an inference in which we have a vector of observable summary
statistics \vecc\ that we are using to constrain a set of parameters of
interest \vecq.  There is a model $\hatc(\vecq,\vecn)$ for the
observables which involves the parameters of interest, but also a
vector \vecn\ of nuisance parameters.  
We wish to characterize the Bayesian posterior probability
\begin{equation}
  p(\vecq | \vecc) \propto \int dn\, \likeli(\vecc | \vecq, \vecn) p(\vecq) p(\vecn),
\label{eq:posterior}
\end{equation}
where $\likeli(\vecc | \vecq, \vecn)$ is a known likelihood function
of the data, and $p(\vecq)$ and $p(\vecn)$ are priors on the
parameters.  This posterior is complex enough that it requires
approximation by the output of a Markov chain (MC) wandering across the space $(\vecq,\vecn).$

@@

A concrete example, which motivated this development,
is when the observable data \vecc\ are the binned 2-point correlation functions of
cosmic fields derived from a catalog of galaxies; the parameters of
interest are cosmological quantities such as $\Omega_m, \sigma_8,$
etc.; and the nuisance parameters $\vecn$
include the coefficients of some linear expansion of the redshift distributions $n(z)$ of the galaxies being observed:
\begin{equation}
  n(z) = \sum_{k=1}^{M} n_k b_k(z).
  \label{eq:nzbasis}
\end{equation}
The $b_k$ are a set of predetermined basis functions for the redshift
distribution.  In our case there are 10 distinct galaxy populations to
be characterized, each with $n(z)$ tabulated at $\approx100$ values, leading to $O(1000)$ parameters $n_k$ to be considered.

@@

Facing this problem for the cosmological analyses of the 3-year data (Y3)
of the \textit{Dark Energy Survey} (DES),
\citet{hyperrank} devised a scheme whereby
the samples of \vecn\ are placed in a grid within some $M$-dimensional
unit hypercube $\mathcal{H}$.  The coordinates \vecu\ within the hypercube
are considered the compressed parameters of $n(z),$ and the
decompression function
$\hat{\vecn}(\vecu)$ outputs the $\vecn_\alpha$ sample at the nearest grid point to
  \vecu.  This solves the problem of creating a continuous \vecu\
  domain, and maintains the equal prior probability of each \vecn\
  sample, but the function \emph{output,} and the resultant likelihood
  function of \vecu, are discontinuous.
Various strategies are proposed by \citet{hyperrank} to assign the $\vecn_\alpha$ to the grid points in
$\mathcal{H}$ based on summary statistics, to reduce the
discontinuities---but the function is never smooth.
As a consequence, many MC samplers become quite inefficient in
sampling of the cosmological posterior.  In particular, samplers such as \textsc{MultiNest} that assume continuity are rendered nearly non-functional.  As a result, the Y3 cosmological priors could not be evaluated with this method.  Instead, the \vecn\ samples were not used, and an \textit{ad hoc} $p(\vecn)$ was adopted which allowed only shifts and dilations of the mean $n(z)$ of the \vecn\ samples.

@@

We describe here an approach that is algebraically similar to that of
\citet{hans}, but does not require 2nd condition of Gaussianity for
the nuisance prior, and is likely to be a good approximation even when
the first and third conditions are not grossly violated.  Our approach is to seek a linear compression of \vecn\ into a lower-dimensional set of parameters \vecu\ that projects away variations in \vecn\ that do not influence the likelihood $\likeli.$  Standard density estimators can then be applied to the \vecu\ values implied by the known \vecn\ samples to yield a prior $p(\vecu)$ that can be used for the MC chain of the cosmological posterior.  The model \hatc, and hence $\likeli,$ will be continuous over this low-dimensional \vecu\ space, and marginalization over \vecu\ will yield posterior probabilities very close to marginalization over the original \vecn.

@@

Since $\matX$ is idempotent, we can write
\begin{align}
  \matX & = V_X \proj_M V_X^T, \\
  \matY \equiv \matI-\matX & = V_X \proj_{-M} V_X^T,
\end{align}
where $V_X$ is unitary and the projection matrix $\proj_M$ is defined as
\begin{equation}
  \left(\proj_M\right)_{ij} \equiv
\begin{cases}
                                            1,  &  i=j\le M \\
                                            0,  & \text{otherwise}
\end{cases}
\end{equation}
and $\proj_{-M}=\matI-\proj_M.$  For a chosen rank $M$ of the
transformation matrix $X$, our task becomes to identify the
eigenvectors $V_X$ that minimize
\begin{align}
  \left\langle \chi^2\right\rangle & = \frac{1}{N_\alpha} \sum_\alpha
  \left[ \matY \vecn_\alpha\right]^T \matF  \left[\matY
                                     \vecn_\alpha\right] \\
       & = \trace \left[ \covm_n V_X \proj_{-M} V_X^T \matF V_X
         \proj_{-M} V_X^T \right], \\
  C_n & \equiv \left\langle \vecn \vecn^T \right\rangle.
\end{align}
This optimization is easier if we transform first the systematic variables to
$\vecn^\prime =\matT\vecn$ such that $\covm_{n^\prime}=\matI$, \ie\
make the elements of $\vecn$ uncorrelated and unit-variance.  This
is accomplished by finding the eigensystem $\covm_n=\matV_n \matLam_n
\matV_n^T$ and setting $\matT = \matLam_n^{-1/2} \matV_n^T$.  With
this transformation, we are now seeking a different unitary matrix
$\matV_{X^\prime}$ that minimizes
\begin{align}
  \left\langle \chi^2\right\rangle & = \trace \left[ \matI
    V_{X^\prime} \proj_{-M} V_{X^\prime}^T \left[ (T^{-1})^T \matF
      T^{-1} \right] V_{X^\prime} 
    \proj_{-M} V_{X^\prime}^T \right] \\
  & = \trace \left[ \proj_{-M} V_{X^\prime}^T \matV_G \matLam_G \matV_G^T
    V_{X^\prime} \proj_{-M} \right],
  \label{eq:mintr}\\
  \matG \equiv \left(T^{-1}\right)^T \matF   T^{-1} & = \matLam_n^{1/2} \matV_n^T
  \matF \matV_n \matLam_n^{1/2} = \matV_G \matLam_G \matV_G^T.
  \label{eq:defG}
\end{align}
The last line defines a transformed Fisher matrix $\matG$ and its
(non-negative) eigensystem.  Examination of \eqq{eq:mintr} reveals that this quantity
is minimized if the elements of the unitary matrices satisfy
$\matV_{X^\prime}^T \matV_G = \matI \; \Rightarrow \; \matV_{X^\prime} = \matV_G,$ and the eigensystem of $G$ is placed
in order of decreasing eigenvalue $\lambda^G_i.$ The
elements surviving the projection $P_{-M}$ yield
\begin{equation}
  \left\langle \chi^2\right\rangle = \sum_{i>M} \lambda_i^G.
  \label{eq:chiresid}
\end{equation}
In other words each eigenvalue of the matrix $\matG$ in \eqq{eq:defG}
gives the contribution to $\left\langle\chi^2\right\rangle$ of one
projection (mode) of $\vecn.$

@@

The orange violins in Figure~\ref{fig:violins} show the distributions
of the $n(z)$ values that are generated by drawing $\widetilde{\vecu}$
values from a unit Gaussian.  It is clear now that projecting away the
unobservable fluctuations in $n(z)$ has substantially reduced the
variance of the function.  To check whether we have achieved our goal
of leaving the \emph{observable} consequences of the $n(z)$ variation
unchanged, we calculate the distribution of
\begin{equation}
  \chi^2 =  \left[\hatc(\vecq,\vecn) - \hatc(\vecq,\vecn_0) \right]^T
                                           \covm_c^{-1}  \left[ \hatc(\vecq,\vecn) - \hatc(\vecq,\vecn_0)\right]
\label{eq:chihist}
\end{equation}
for the cases (1) the $\vecn$ are drawn from the input samples, and
(2) are generated using the procedure defined above.  This $\chi^2$
measures the deviation of the model from that implied by the mean
vector $\vecn_0$ at some chosen nominal value of cosmological
parameters $\vecq$ (we also fix the other nuisance parameters of the
DES model for this test).

Figure~\ref{fig:chihist} shows the results: the $\chi^2$ distribution
for the input samples is indistinguishable from that of the samples
generated by our compressed, normalized representation.
The Figure also shows the $\chi^2$ distribution resulting from samples
generated by the method used in the DES Y3 analysis \citep{y3pz}.  In
that case, the $n(z)$ for each galaxy sample was given an \textit{ad
  hoc} variation of the form
\begin{equation}
  n(z) = n_0\left(\frac{z-\Delta\,z}{s}\right),
\end{equation}
with $\Delta z$ and $s$ being ``shift'' and ``stretch'' parameters.
Separable Gaussian priors were assigned to $\Delta z$ and $s$ with
means of 0 and 1, respectively, and standard deviations that equaled
the RMS variation in the mean and width of the input $n(z)$ samples.
The shift-stretch model essentially compresses the samples of $n(z)$
into two parameters, their mean and standard deviation, and executes
an \textit{ad hoc} reconstruction based on those parameters.
In the Figure we can see that the shift-stretch model does not in fact
reproduce the size of the deviations in the $\vecc$ observables that
is implied by the original samples.
